---
title: "STAT 530 HW1"
author: "Md Muhtasim Billah"
date: "1/28/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 1.	

Write TRUE or FALSE with reasons for the following:

# Answer

**a.**	Number on the jerseys of Basketball players is a numerical variable.

**FALSE.**

\underline{Reason:} The number on the jerseys of basketball players are onlys used to identify the players. It has neither any mathematical meaning nor it can be manipulated, rather it has categorical value. Thus, it can be considered as a categorical variable, not numerical.\newline

**b.**	If y follows normal with mean 4 and variance 1, y^2 follows a chi-square distribution with 1 df.

**TRUE.**

\underline{Reason:}
Though chi-square distribution requires standard normal distribution (mean=0 and variance=1), not being so won't affect the degree of freedom. So, for the given case, the df will be 1 but it will be a non-central chi-squre distribution.\newline

**c.**	If y1 is a normal with mean 0 and variance 1, and y2 follows chi-square with n-1 degrees of freedom then y1/sqrt(y2/n-1) follows a F distribution with (n-1) degrees of freedom.

**FALSE.**

\underline{Reason:}
It will be a student's t-distribution.\newline

**d.**	If y3 follows a F distribution then square root of y3 follows t.

**FALSE.**

\underline{Reason:}
It will not necessarily be a t-distribution.\newline



# Question 2.	

Consider the data set on lead concentration and mortality of midge flies given in our class website names Hw1-2020.xls.  or Hw1-2020.csv (same data set in two formats). Answer the questions based on this data set:


a.	Why is mortality considered the response variable?  Why is lead concentration the explanatory variable?

b.	Plot the relationship.  What do you see from the plot?

c.	Determine least squares’ equation that can be used for predicting mortality.

d.	Is the slope significantly greater than 0.  Use alpha = .05.

e.	Use your model to predict the mortality rate percentage when lead concentration is 225 ppm.   

f.	Provide 95% confidence intervals for e.

g.	Provide 95% prediction intervals for (e)

h.	Comment on your findings in (f) and (g).

# Answer:

## a.

Lead concentration is considered the explanatory variable because it can be easily measured and the mortality rate is directly dependent on the lead concentration.

The mortality rate is the response variable because it is dependent directly on the explanatory varible- the lead concentration and we can make an inference about it based on the explanatory variable.

## b.

Plotting the relationship between lead concentration and mortality rate in R.

```{r plot}
setwd("/Users/muhtasim/Desktop/STAT530/HWs")
getwd()
data1=read.csv("HW_1_2020.csv",header=TRUE,sep=",",na.strings=" ")
head(data1)
leadconc=data1[,1]
mortality=data1[,2]
plot(leadconc,mortality)
```
\newline
From the plot, it is evident that the mortality rate almost linearly increases with lead concentration.

## c.

The model for lest square method is:
$$Y=\beta_0+\beta_1x+\varepsilon$$
Where, $\beta_0$ is the intercept, $\beta_1$ is the slope and $\varepsilon$ is the error.\newline
For determining the least squares' equation, we can use the linear regression model in R.
```{r }
fit=lm(mortality~leadconc, data=data1)
coefficients(fit)
```
From R, we find that the intercept is $\beta_0=61.24$ and the positive slope is $\beta_1=0.03$. So, the least squares' equation will be,
$$Y=61.24+0.03x+\varepsilon$$

## d.

To see if the slope is significantly greater than 0 or not, we can run a summary in R for the linear regression model.

```{r }
summary(fit)
```
Looking at the significance codes in the summary output, it can be concluded that the slope is not significantly greater than 0 (using $\alpha=0.05$)


## e.

Using the linear model, the mortality rate percentage can be predicted for lead concentration of 225 ppm. 
```{r }
y=mortality
x=leadconc
#prediction of mortality for leadconc of 225 ppm
new=data.frame(x = seq(225,0.05))
predict(lm(y ~ x), new)
```

From the generated data, the predicted mortality rate at lead concentration of 225 ppm is 61.27%. 

## f, g.

The plot is given below for the pediction with 95% confidence intervals and 95% prediction intervals.
```{r }
pred.w.plim=predict(lm(y ~ x), new, interval = "prediction")
pred.w.clim=predict(lm(y ~ x), new, interval = "confidence")
#prediction plots
matplot(new$x, cbind(pred.w.clim, pred.w.plim[,-1]),  type = "l", ylab = "predicted y")
```

## h.

The plot specifies the computation of confidence or prediction (tolerance) intervals at the specified level (95%). It is also referred to as narrow vs. wide intervals. This provides an idea about how good the prediction is.

# Question 3.	
Why is the regression estimates called the LEAST SQUARE estimates. Comment.

# Answer:

Regression estimates can largely be divided into two categories - least square method and maximum likelihood method. The model for estimating the parameters for both the methods is following:
$$Y=\beta_0+\beta_1x+\varepsilon$$
The error estimate from this equation is,
$$\varepsilon=Y-(\beta_0+\beta_1x)$$
But, $\sum\varepsilon$ will always be zero, so minimization of this quantity doesn't make any sense. So, instead we find the summation of the square of the error given by the following expression.

$$\sum\varepsilon^2=\sum\{Y-(\beta_0+\beta_1x)\}^2$$
That is why the regression estimates are called the LEAST SQUARE estimates. This entitiy is also visible in case of maximum likelihood method. The likelihood function is given as below.
$$L(\beta_0,\beta_1)=\prod_{i = 1}^{n}\frac{1}{\sqrt{2\pi}}exp\left[\frac{\{y_i-(\beta_0+\beta_1x_i)\}^2}{2\sigma^2}\right]$$
On the right side of the expression, we can see the same entity which is the least squared error estimate and that is why the name LEAST SQUARE is used for the regression estimate.


# Question 4.
Give an example from your discipline where you think linear regression would be appropriate.

# Answer:

Linear regression can be applied in so many scientific fields. I am a mechanical engineering major and heat conduction through material is a common phenomena. As it turns out, with increased thermal conductivity of a material, the heat transfer rate is increased alsmost linearly and linear regression can be fruitfully used to predict the heat transfer rate in a material.


